How to Avoid OOTA Without Really Trying

### 2025-01-15 paper/rebuttal.txt

First, we thank all the reviewers for their time, their careful reviews,
and their thoughtful feedback.

Several reviewers suggested citations that we are happy to add.

Other reviewers asked how we handle various litmus tests.  Our submission
describes how we rely on a compiler to translate the litmus test to object
code, then relate dependencies in that object code to the corresponding
source code.  Section 1.2 ("Prior Work") notes that we are not trying
to identify when OOTA cycles can form, but rather that OOTA cycles
cannot form.

Several reviewers seemed uncomfortable with informal proofs.  However,
ASPLOS has published papers with informal proofs, so we suspect that the
real discomfort was augmenting the C++ abstract machine with hardware
properties.  20 years of equivocal progress is more than sufficient to
justify trying a new approach.  We are not advocating the old approaches
be abandoned.  In fact, should someone succeed with those approaches,
we will endeavor to be the first to offer congratulations.

Others were uncomfortable with our reliance on object code, which
varies across systems.  However, these systems are underpinned by laws
of physics, namely causality, the finite speed of light, and the atomic
nature of matter.  Our definition of OOTA cycles as cycles in (rfe U
sdep) permits us to examine rfe and sdep separately.  These laws of
physics guarantee that each rfe link goes forward in time.  Because an
OOTA cycle must return to the starting time, there must be at least
one sdep link that goes backwards in time.  Correct compilers guarantee
that sdep links correspond to sequences of machine instructions, which
cannot execute backwards in time, even given speculative execution.
Thus, the OOTA cycle cannot form.

Please note that correct implementations are not required to break
non-semantic dependencies.  An unbroken non-semantic dependency will
correspond to a sequence of machine instructions that cannot go backwards
in time, and thus cannot form an OOTA cycle.

Reviewer 72F asks about three litmus tests in Section 5 of the
submission's citation [5], as do a couple of other reviewers:

o	The first example has a dependency only if the implementation
	does not hoist the i++ out of the “if” statement, and this
	is the implementation’s choice, similar to the example in
	Section 2.2.5 of the submission.

o	The second example is handled by examining f() and the functions
	that it invokes, along with the implementation’s optimization
	choices.  Note that at one extreme, f() might ignore its
	parameter and at the other extreme it might simply return
	its parameter.  Note further that f() might be a pointer to a
	function, and that function might come from a dynamically linked
	library that has not yet been written at the time the call to
	f() is compiled.  Although the submission does not contain
	this particular example, all of these issues are addressed by
	the submission’s Section 3, which discusses executions (as
	opposed to source-code syntactic dependencies).

o	An instructive elaboration of the third example is discussed in
	Section 2.2.1 of the submission.

Reviewer 72F also asks if our approach implicitly enforces the
load-to-store ordering advocated by Boehm and Demsky.  The answer is
"no".  First, although instructions commit in some order, they do not
necessarily commit in program order, which on some systems allows prior
loads to be reordered with later stores.  Even on hardware that does
commit instructions in program order, there is no guarantee that this
commit order will constrain the visibility of the corresponding memory
references, for example, due to the effects of hardware store buffers.

Reviewer 72F also asks (in two questions) if freezing of nondeterministic
choices into machine code is realistic.  The answer is "yes", but on the
per-execution basis used in our submission.  For example, specialization
optimizations might cause a given function to be compiled differently
for different combinations of arguments.  But any given execution would
correspond to a particular combination of arguments, and that execution
would be frozen into machine code.  Please note that programs that
modify their own code are outside of the standard, and thus out of our
submission's scope.

We especially thank Reviewer 72F for noting (correctly!) that Section 7.4
("Merging Quasi-Volatile Loads") poses a challenge, as this helped refine
our thinking. The refinement is to note that this challenge is not to our
technique, but rather to those implementing optimizations.  Although we
believe (one of us strongly, the rest less so) that the "optimization"
described in this section is in fact a pessimization that also degrades
ease of use, the question of whether or not this optmization is legitimate
will be decided by community processes, not by fiat and not by pure
mathematics.  Nevertheless, should an abstract-machine mathematical
treatment of OOTA appear, we would expect that this mathematics would
have a prominent seat at the community-process table.  Either way,
if this optimization is rejected the litmus test in Section 7.4 has a
semantic dependency, otherwise it does not.  And correct implementations
will reflect that decision, though again, optimizations are permitted
to refrain from breaking non-semantic dependencies.

Reviewer 72F suggested that rf links must impose restrictions.
Please note our Section 2.1 ("OOTA: rf vs. rfe") where we show that
those restrictions can be extremely loose, hence our use of rfe.
But note also that C++ modification order provides very weak guarantees,
in fact, a later-in-time store might precede an earlier-in-time store
in modification order, even on x86.  Memory models are at best loosely
connected to the passage of time.

Reviwer 72F also questions how the original C++ code can be irrelevant.
The C++ code is relevant, but only at the source-code level and to the
connections between the source-code and object-code levels.  However,
for analysis confined to the object-code level, the origial C++ code
can indeed be irrelevant.

Reviewer 72F asks what we mean by "these models are atemporal".  We
mean that typical memory models do not make use of the passage of
physical time.  The provide partial orderings, but as noted earlier,
the modification order can in fact go backwards in time.

Responses to individual reviewer feedback items is in the accompanying
PDF.

### 2025-01-15 paper/rebuttal.txt

First, we thank all the reviewers for their careful reviews and thoughtful
feedback.

We are happy to add the suggested citations.

There was discomfort with informal proofs.  However, ASPLOS has published
such papers, so we suspect that the real discomfort was augmenting the
C++ abstract machine with hardware properties.  20 years of equivocal
progress justifies trying a new approach.  We are not advocating the old
approaches be abandoned.  In fact, we will endeavor to be the first to
offer congratulations should someone succeed.

Various litmus tests were put forward.  Our submission describes how
we rely on a compiler to translate the litmus test to object code, then
relate dependencies in that object code to the corresponding source code.
Section 1.2 ("Prior Work") notes that we are not trying to identify when
OOTA cycles can form, but rather show that OOTA cycles cannot form.

There was also discomfort with use of object code, which varies across
systems.  However, these systems are underpinned by laws of physics,
namely causality, the finite speed of light, and the atomic nature
of matter.  Our definition of OOTA cycles as cycles in (rfe U sdep)
permits us to examine rfe and sdep separately, noting that each rfe link
goes forward in time.  Because an OOTA cycle must return to the starting
time, there must be at least one sdep link that goes backwards in time.
Correct compilers guarantee that sdep links correspond to sequences of
machine instructions, which cannot execute backwards in time, even given
speculative execution.  Thus, the OOTA cycle cannot form.

Please note that correct implementations are not required to break
non-semantic dependencies.  Regardless, a non-semantic dependency cannot
form an OOTA cycle.

@@

Reviewer 72F asks about three litmus tests in Section 5 of the
submission's citation [5], as do a couple of other reviewers:

o	The first example has a dependency only if the implementation
	does not hoist the i++ out of the “if” statement, and this is
	the implementation’s choice, similar to the example in Section
	2.2.5 of the submission.  See Sections 2.2.7 and 2.2.8 for more
	aggressive examples.

o	The second example is handled by examining f() and the
	functions that it invokes, along with the implementation’s
	optimization choices.  This example is yet another motivation
	for the submission’s Section 3, which discusses executions as
	opposed to source-code syntactic dependencies.

@@

o	An instructive elaboration of the third example is discussed in
	Section 2.2.1 of the submission.

Reviewer 72F also asks if we implicitly enforce the load-to-store
ordering advocated by Boehm and Demsky.  We do not.  Instructions do
not necessarily commit in program order.  Furthermore, commit order
might not constrain the order of the corresponding memory references.

Reviewer 72F also asks (in two questions) if freezing of nondeterministic
choices into machine code is realistic.  The answer is "yes", but on the
per-execution basis used in our submission.  For example, specialization
optimizations might cause a given function to be compiled differently
for different combinations of arguments.  But any given execution would
correspond to a particular combination of arguments, and that execution
would be frozen into machine code.  Please note that self-modifying
programs are outside of the standard, and thus out of our scope.

We thank Reviewer 72F for noting that Section 7.4 ("Merging Quasi-Volatile
Loads") poses a challenge.  However, please note that this challenge
is not to our technique, due to our focus on object-code executions.
Community processes will decide whether or not this section's optimization
is legitimate, and these processes' outcomes will be implemented in
compilers, thus being automatically handled by our technique.  The
challenge is instead to OOTA analysis techniques not accounting for these
community processes.  Nevertheless, any abstract-machine mathematical
OOTA solution should have a prominent seat at the community-process table.

Reviewer 72F suggested that rf links must impose restrictions.
Please note our Section 2.1 ("OOTA: rf vs. rfe") where we show that
those restrictions can be extremely loose, hence our use of rfe.
But note also that C++ modification order provides very weak guarantees,
in fact, a later-in-time store might precede an earlier-in-time store
in modification order, even on x86.

@@

Reviwer 72F also questions how the original C++ code can be irrelevant.
The C++ code is relevant, but only at the source-code level and to the
connections between the source-code and object-code levels.  However,
for analysis confined to the object-code level, the origial C++ code
can indeed be irrelevant.

Reviewer 72F asks what we mean by "these models are atemporal".
We mean that typical memory models do not make use of the passage of
physical time.  The provide partial orderings, but as noted earlier,
some orderings can go backwards in time.

### 2025-01-15 paper/rebuttal.txt

First, we thank all the reviewers for their careful reviews and thoughtful
feedback.

Responses to individual reviewer feedback items may be found in the
accompanying PDF.

@@

We are happy to add the suggested citations.

There was discomfort with informal proofs.  However, our proofs are
straightforward and ASPLOS has published such proofs.  We therefore
suspect that the real discomfort was augmenting the C++ abstract machine
with hardware properties.  20 years of equivocal progress justifies trying
a new approach.  We are not advocating the old approaches be abandoned.
In fact, we will endeavor to be the first to offer congratulations should
someone succeed.

@@

Various litmus tests were put forward.  Our submission describes how
we rely on a compiler to translate the litmus test to object code, then
relate dependencies in that object code to the corresponding source code.
Section 1.2 ("Prior Work") notes that we are not trying to identify when
OOTA cycles can form, but rather show that OOTA cycles cannot form.

There was also discomfort with use of object code, which varies across
systems.  However, these systems are underpinned by laws of physics,
namely causality, the finite speed of light, and the atomic nature
of matter.  Our definition of OOTA cycles as cycles in (rfe U sdep)
permits us to examine rfe and sdep separately, noting that each rfe link
goes forward in time.  Because an OOTA cycle must return to the starting
time, there must be at least one sdep link that goes backwards in time.
Correct compilers guarantee that sdep links correspond to sequences of
machine instructions, which cannot execute backwards in time, even given
speculative execution.  Thus, the OOTA cycle cannot form.

Please note that correct implementations are not required to break
non-semantic dependencies.  Regardless, a non-semantic dependency cannot
form an OOTA cycle.

Responses specifically requested by Reviewer 72F for the rebuttal:

How do we handle the three litmus tests in Section 5 of citation [5]?

@@

o	The first example has a dependency only if the implementation
	does not hoist the i++ out of the “if” statement, and this is
	the implementation’s choice, similar to the example in Section
	2.2.5 of the submission.  See Sections 2.2.7 and 2.2.8 for more
	aggressive examples.

o	The second example is handled by examining f() and the
	functions that it invokes, along with the implementation’s
	optimization choices.  This example is yet another motivation
	for the submission’s Section 3, which discusses executions as
	opposed to source-code syntactic dependencies.

o	An instructive elaboration of the third example is discussed in
	Section 2.2.1 of the submission.

Do we implicitly enforce the load-to-store ordering advocated by Boehm
and Demsky?  (Two questions.)  We do not.  Instructions do not necessarily
commit in program order.  Furthermore, commit order might not constrain
the order of the corresponding memory references.

Is freezing of nondeterministic choices into machine code realistic?
Yes, but on the per-execution basis used in our submission.  For example,
specialization optimizations might cause a given function to be compiled
differently for different combinations of arguments.  But any given
execution would correspond to a particular combination of arguments,
and that execution would be frozen into machine code.

@@

We thank Reviewer 72F for noting that Section 7.4 ("Merging Quasi-Volatile
Loads") poses a challenge.  However, please note that this challenge
is not to our technique, due to our focus on object-code executions.
Community processes will decide whether or not this section's optimization
is legitimate, and these processes' outcomes will be implemented in
compilers, thus being automatically handled by our technique.  The
challenge is instead to OOTA analysis techniques not accounting for these
community processes.  Nevertheless, any abstract-machine mathematical
OOTA solution should have a prominent seat at the community-process table.

Don't rf links must impose restrictions?  Yes, they do in the C++ memory
mode, but only when they involve release-stores and acquire-loads or
consume-loads.  These restrictions do not apply to OOTA cycles, which
involve only relaxed accesses.  Note further that C++ modification
order provides very weak guarantees, in fact, a later-in-time store
might precede an earlier-in-time store in modification order, even on x86.

How the original C++ code can be irrelevant?  The C++ code is relevant,
but only at the source-code level and to the connections between the
source-code and object-code levels.  However, for analysis confined to
the object-code level, the original C++ code can indeed be irrelevant.
For example, if a program in C++ and another in Rust generated the same
machine code, the behavior of that machine code would be independent of
the source language.

What is meant by "these models are atemporal"?  We mean that typical
memory models do not make use of the passage of physical time.
The provide partial orderings, but as noted earlier, some orderings can
go backwards in time.

### 2025-01-16 paper/rebuttal-mods.txt

The "o" items assume a venue (such as a tech report) that have relaxed
page limits.

O	Detailed feedback from Review #72C.

O	s/We would be happy to change "explicitly prohibits"/"explicitly recommends against"/
	This is prior to the quote from the C++ standard.

O	s/Workers/Researcher/ in Section 1.2.

O	Add citations to Section 1.2 ("Prior Work").

O	Be less condescending towards Mark Batty.  Maybe
	s/is persisting with/is working on/.

o	Add a Section 1.3 ("Our Approach"?) that summarizes the
	rfe/sdep/non-sdep-dep approach and emphasizes proving
	OOTA cannot happen as opposed to an exact delineation
	between sdep and non-sdep.

o	Add citations to later sections that discuss similar litmus tests.

o	Consider adding a 2.2.x section that discusses the difficulties
	inherent in computing sdep for function calls, especially when
	the function will be part of a dynamic library that has not yet
	been written.

O	In section 6.1, define "witness" informally as "another execution
	in which the store acts differently".

O	On page 9, "While it possible".  [ Since fixed by Alan. ]

O	In Section 7.4 ("Merging Quasi-Volatile Loads"), change focus
	from intuitive notions to a community-process approach for
	determining what optimizations are legitimate.  Math cannot
	handle this.  We avoid the problem by looking at the compiler
	implementation, which is downstream of the community process.

	Also in Section 7.4 change "our intuitive notions" to "one's
	intuitive notions".

o	Possible future work: Relax quasi-volatile semantics to permit
	additional uses of the cmov instruction.

### 2025-01-21 paper/rebuttal-mods.txt

The "o" items assume a venue (such as a tech report) that have relaxed
page limits.

O	Detailed feedback from Review #72C.

o	Early in the paper, consider explicitly stating that our approach
	enables us to avoid the need to precisely delineate between sdep
	and non-sdep dependencies.

@@

o	Add citations to later sections that discuss similar litmus tests.

o	Consider adding a 2.2.x section that discusses the difficulties
	inherent in computing sdep for function calls, especially when
	the function will be part of a dynamic library that has not yet
	been written.

O	In section 6.1, define "witness" informally as "another execution
	in which the store acts differently".

o	Add diagrams illustrating what the various components of sections
	6.1 and 6.3 refer to.

@@

O	On page 9, "While it possible".  [ Since fixed by Alan. ]

O	In Section 7.4 ("Merging Quasi-Volatile Loads"), change focus
	from intuitive notions to a community-process approach for
	determining what optimizations are legitimate.  Math cannot
	handle this.  We avoid the problem by looking at the compiler
	implementation, which is downstream of the community process.

	Also in Section 7.4 change "our intuitive notions" to "one's
	intuitive notions".

o	Add a section before Section 7 ("Issues and Refinements") that
	shows how our approach handles a few well-known examples of OOTA.
	This needs to be in greater detail than in the "Causality Test
	Cases" discussions in P3064R2 ("How to Avoid OOTA Without Really
	Trying").

o	Possible future work: Relax quasi-volatile semantics to permit
	load speculation, for example, use of the cmov instruction in
	cases where the source code conditionally executes the atomic
	load.

### 2025-01-22 paper/rebuttal-mods.txt

The "o" items assume a venue (such as a tech report) that have relaxed
page limits.

O	Detailed feedback from Review #72C.

o	Early in the paper, consider explicitly stating that after 20
	years it appears that the problem of characterizing semantic
	dependency is too difficult to solve in the context provided by
	the C++ standard (i.e., the abstract machine) and therefore we
	expand the context by considering actual machine-level executions.

@@

O	s/We would be happy to change "explicitly prohibits"/"explicitly recommends against"/
	This is prior to the quote from the C++ standard.

O	s/Workers/Researcher/ in Section 1.2.

O	Add citations to Section 1.2 ("Prior Work").

	Add a forward reference to Section 4.2 ("Global Optimization
	Can Destroy Dependencies") and the fact that this defeats the
	approach of ordering relaxed loads before relaxed stores.

@@

o	Add a Section 1.3 ("Our Approach"?) that summarizes the
	rfe/sdep/non-sdep-dep approach and emphasizes proving
	OOTA cannot happen as opposed to an exact delineation
	between sdep and non-sdep.

o	Add citations to later sections that discuss similar litmus tests.

o	Consider adding a 2.2.x section that discusses the difficulties
	inherent in computing sdep for function calls, especially when
	the function will be part of a dynamic library that has not yet
	been written.

o	In Section 4.2 ("Global Optimization Can Destroy Dependencies"),
	note that this optimization breaks the approach of ordering
	relaxed loads before relaxed stores.

O	In section 6.1, define "witness" informally as "another execution
	in which the load(s) obtain a different value but all else is
	the same as far as possible, and yet the store acts differently".

### 2025-02-28 paper/rebuttal-mods.txt

o	Add a Section 1.3 ("Our Approach"?) that summarizes the
	rfe/sdep/non-sdep-dep approach and emphasizes proving
	OOTA cannot happen as opposed to an exact delineation
	between sdep and non-sdep.

o	Add citations to later sections that discuss similar litmus tests.

o	Consider adding a 2.2.x section that discusses the difficulties
	inherent in computing sdep for function calls, especially when
	the function will be part of a dynamic library that has not yet
	been written.

o	In section 2.2 where we start to talk about semantic dependencies
	in detail, we could mention syntactic dependencies too.  Explain
	what they are and how they differ from semantic dependencies.
	Then mention that some work in the literature (such as parts of
	[5]) is not always careful to distinguish clearly between the
	two forms.  Maybe have an entire subsection on this topic.

### 2025-02-28 paper/rebuttal-mods.txt

The "o" items assume a venue (such as a tech report) that have relaxed
page limits.

O	Detailed feedback from Review #72C.

o	Early in the paper, consider explicitly stating that after 20
	years it appears that the problem of characterizing semantic
	dependency is too difficult to solve in the context provided by
	the C++ standard (i.e., the abstract machine) and therefore we
	expand the context by considering actual machine-level executions.

x	s/We would be happy to change "explicitly prohibits"/"explicitly recommends against"/
	This is prior to the quote from the C++ standard.

x	s/Workers/Researcher/ in Section 1.2.

@@

O	Add citations to Section 1.2 ("Prior Work").

x	Add a forward reference to Section 4.2 ("Global Optimization
	Can Destroy Dependencies") and the fact that this defeats the
	approach of ordering relaxed loads before relaxed stores.

x	Be less condescending towards Mark Batty.  Maybe
	s/is persisting with/is working on/.

@@

o	Add a Section 1.3 ("Our Approach"?) that summarizes the
	rfe/sdep/non-sdep-dep approach and emphasizes proving
	OOTA cannot happen as opposed to an exact delineation
	between sdep and non-sdep.

o	Add citations to later sections that discuss litmus tests
	that are similar to those in the literature.

@@

o	Consider adding a 2.2.x section that discusses the difficulties
	inherent in computing sdep for function calls, especially when
	the function will be part of a dynamic library that has not yet
	been written.

o	In section 2.2 where we start to talk about semantic dependencies
	in detail, we could mention syntactic dependencies too.  Explain
	what they are and how they differ from semantic dependencies.
	Then mention that some work in the literature (such as parts of
	[5]) is not always careful to distinguish clearly between the
	two forms.  Maybe have an entire subsection on this topic.

x	In Section 4.2 ("Global Optimization Can Destroy Dependencies"),
	note that this optimization breaks the approach of ordering
	relaxed loads before relaxed stores.

x	In section 6.1, define "witness" informally as "another execution
	in which the load(s) obtain a different value but all else is
	the same as far as possible, and yet the store acts differently".

### 2025-02-28 paper/ootavacuous.tex
ootavacuous: De-anonymize
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Bibliography update and handling of trivial rebuttal items
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

% Blind review process.
% \author{
% Alan Stern\\{\small \email{stern@rowland.harvard.edu}} \and
% Paul E.~McKenney\\{\small \email{paulmck@kernel.org}} \and
% Michael Wong\\{\small \email{fraggamuffin@gmail.com}} \and
% Maged Michael\\{\small \email{maged.michael@gmail.com}}
% }

@@

Because nothing in the C++ memory model rules out such OOTA cycles
(as indicated by the ``sometimes satisfied'' in the figure), the C++
standard explicitly prohibits them in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
\begin{quote}
	Implementations should ensure that no “out-of-thin-air” values
	are computed that circularly depend on their own computation.
\end{quote}
This prohibition prevents misapplication of the as-if rule in oracular
C++ implementations.

@@

All OOTA workers owe a debt of gratitude to the foundational work led
by William Pugh, represented by the infamous Java-language
``Causality Test Cases''.\footnote{
	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
% %%% Eventually cite P3064.

@@

Some workers recommend avoiding OOTA by forcing prior relaxed
loads to be ordered before subsequent relaxed
stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352},
but this can require executing real
instructions~\cite[Section 7.1]{Maranget2012TutorialARMPower},
consuming real time and real electrical power to solve a strictly
theoretical problem.
% This might have been acceptable in the 1960s of some of the authors'
% youths, but it is now the year 2024.

Other workers recommend various procedures to identify and avoid OOTA
cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
but none of these have been accepted by C++ implementers.
As of 2024, Mark Batty is persisting with modular relaxed dependencies,
which may have the added ability of checking compiler
optimizations~\cite{JayRichards2024SymbolicMrder}.

@@

Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
page~\pageref{lst:Simple OOTA},
in which T0 loads the value of \co{x} and stores it in \co{y} while T1
does the reverse.
A sufficiently perverse globally optimizing loose C++ compiler might
transform that program to the following before translating it into
machine code:
% Constructed by hand.
\begin{quote}
\small
\begin{tabular}{r|l|l}
% & \multicolumn{1}{c}{T0} & \multicolumn{1}{c}{T1} \\
% \hline
1. &
\texttt{~~int r1 = 42} &
	\texttt{~~int r2 = 42} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~x = r2} \\
\end{tabular}
\end{quote}
The loads previously on line~1 have been replaced by constants.
Such a transformation complies with the loose C++ standard,
even though the resulting executable file would produce an unintuitive
OOTA outcome every time it runs!

@@

>>>>

\author{
Alan Stern (stern@rowland.harvard.edu) \and
Paul E.~McKenney (paulmck@kernel.org) \and
Michael Wong (fraggamuffin@gmail.com)
}

@@

Because nothing in the C++ memory model rules out such OOTA cycles
(as indicated by the ``sometimes satisfied'' in the figure), the C++
standard explicitly recommends against them in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
\begin{quote}
	Implementations should ensure that no “out-of-thin-air” values
	are computed that circularly depend on their own computation.
\end{quote}
This prohibition prevents misapplication of the as-if rule in oracular
C++ implementations.

@@

All OOTA researchers owe a debt of gratitude to the foundational work led
by William Pugh, represented by the infamous Java-language
``Causality Test Cases''.\footnote{
	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
% %%% Eventually cite P3064.

@@

Some workers recommend avoiding OOTA by forcing prior relaxed
loads to be ordered before subsequent relaxed
stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352},
but this can require executing real
instructions~\cite[Section 7.1]{Maranget2012TutorialARMPower},
consuming real time and real electrical power to solve a strictly
theoretical problem.
% This might have been acceptable in the 1960s of some of the authors'
% youths, but it is now the year 2024.
Furthermore, as discussed in
Section~\ref{sec:Global Optimization Can Destroy Dependencies},
cross-thread optimizations can defeat this ordering constraint.

Other workers recommend various procedures to identify and avoid OOTA
cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
but none of these have been accepted by C++ implementers.
As of 2024, Mark Batty is working on modular relaxed dependencies,
which may have the added ability of checking compiler
optimizations~\cite{JayRichards2024SymbolicMrder}.

@@

Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
page~\pageref{lst:Simple OOTA},
in which T0 loads the value of \co{x} and stores it in \co{y} while T1
does the reverse.
A sufficiently perverse globally optimizing loose C++ compiler might
transform that program to the following before translating it into
machine code:
% Constructed by hand.
\begin{quote}
\small
\begin{tabular}{r|l|l}
% & \multicolumn{1}{c}{T0} & \multicolumn{1}{c}{T1} \\
% \hline
1. &
\texttt{~~int r1 = 42} &
	\texttt{~~int r2 = 42} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~x = r2} \\
\end{tabular}
\end{quote}
The loads previously on line~1 have been replaced by constants.
Such a transformation complies with the loose C++ standard,
even though the resulting executable file would produce an unintuitive
OOTA outcome every time it runs, even if relaxed loads are ordered
before relaxed stores!

@@

We informally define a \emph{witness} to be another execution in which
at least one of the loads obtains a different value but all else is the
same as far as possible, and yet the store acts differently.
